---
title: "Multinomial Regression"
author: "Jim Harner"
date: "1/6/2020"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
`sparklyr` requires a `dplyr` compatible back-end to Spark. 
```{r}
library(dplyr, warn.conflicts = FALSE)
library(sparklyr)
# start the sparklyr session
master <- "local"
# master <- "spark://master:7077"
sc <- spark_connect(master)
```

## 6.6 Multinomial Regression

We now look at the classification problem in which there are $k > 2$ groups.

### 6.6.1 Basics

Multinomial regression is a relatively simple extension of logistic regression. We now have $k - 1$ logit transformations expressed linearly in terms of the $X$'s. The last group is used as the denominator in these logits.

We can then compute $P(G = l\, | X = x)$, for $l = 1, 2, \ldots, k - 1$. $P(G = k\, |\, X = x)$ is obtained by subtraction.

### 6.6.2 Multinomial Models

The diabetes data from the Reaven and Miller study has diabetic-related measurements on 145 patients:   
* RelWeight - relative weight  
* GluFast - blood sugar level prior to the glucose tolerance test  
* GluTest - average blood sugar level during the test    
* InsTest - average insulin level during the test  
* SSPG - a measure of how glucose and insulin interact  
* CClass - clinical diagnosis (3=Normal, 2=Chemical Diabetic, 1=Overt Diabetic)  
The dataset is small, but it illustrates binomial models using two levels (by combining Overt and Chemical Diabetics) or multinomial models using three levels.

The `diabetes.csv` file is read into an R data frame and the `CClass` variable is converted from an `int` to a factor with `chr` values.

```{r}
diabetes_df <- read.csv("diabetes.csv", header = TRUE) %>%
  mutate(GluDiff = GluTest - GluFast) %>%
  mutate(CClass = factor(CClass, labels = c("o", "c", "n")))
diabetes_sdf <- copy_to(sc, diabetes_df, "diabetes_sdf")
head(diabetes_sdf)
```
It would be possible to binarize `CClass` by combining `o` and `c` to `d` using:
```
mutate(CClass = recode(CClass, "o" = "d", "c" = "d", "n" = "n"))
```
but we will keep 3 groups.

The `ml_logistic_regression` function accommodates $k > 2$.
```{r}
diabetes_logistic_fit <- diabetes_sdf %>%
  ml_logistic_regression(CClass ~ GluDiff)
diabetes_logistic_fit
```
The output gives the coefficient estimates for each of the three groups. The normal group appears to be quite different than the chemical and overt disbetics.

```{r}
diabetes_logistic_predict <- ml_predict(diabetes_logistic_fit)
diabetes_logistic_predict
```

The f1 performance measure, which combines precision and recall is:
```
ml_multiclass_classification_evaluator(diabetes_logistic_predict, label_col = "CClass",
                            prediction_col = "prediction", metric_name = "f1")
```
whereas the accuracy is:
```
ml_multiclass_classification_evaluator(diabetes_logistic_predict, label_col = "CClass",
                       prediction_col = "prediction", metric_name = "accuracy")
```

```{r}
spark_disconnect(sc)
```


