---
title: "Spark Apply"
author: "Jim Harner"
date: "10/4/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load `sparklyr` and establish the Spark connection.
```{r}
library(dplyr, warn.conflicts = FALSE)
library(sparklyr)

# start the sparklyr session
master <- "local"
# master <- "spark://master:7077"
sc <- spark_connect(master, spark_home = Sys.getenv("SPARK_HOME"),
                    method = c("shell"), app_name = "sparklyr")
```

## 5.4 Spark Apply

`sparklyr` provides support to run arbitrary R code at scale within Spark through the function `spark_apply`. Thus, much of R's functionality can be distributed across an R cluster. Apache Spark, even with Spark Packages, has a limited range of functions available.

`spark_apply` applies an R function to a Spark DataFrame, typically. Spark objects are partitioned so they can be distributed across a cluster. You can use `spark_apply` with the default partitions or you can define your own partitions with the `group_by` argument. Your R function must return another Spark DataFrame. `spark_apply` will run your R function on each partition and output a single Spark DataFrame.

### 5.4.1 Apply an R function to a Spark Object

Let's apply the identify function, `I()`, over a list of numbers we created with the `sdf_len` function.
```{r}
sdf_len(sc, length = 5, repartition = 1) %>%
  spark_apply(function(e) I(e))
```

### 5.4.2 Group By

A common task is to apply your R function to specific groups in your data, e.g., computing a regression model for each group. This is done by specifying a `group_by` argument.

The following example initially counts the number of rows in the `iris` data frame for each species.
```{r}
iris_tbl <- copy_to(sc, iris)
iris_tbl %>%
  spark_apply(nrow, group_by = "Species")
```
Now compute the $R^2$ for a linear model for each species.
```{r}
iris_tbl %>%
  spark_apply(
    function(e) summary(lm(Petal_Length ~ Petal_Width, e))$r.squared,
    names = "r.squared",
    group_by = "Species")
```

### 5.4.3 Distributed Packages

With `spark_apply` you can use nearly any R package inside Spark.

As an example, we use the `broom` package to create a tidy data frame from the linear regression output.
```{r}
spark_apply(
  iris_tbl,
  function(e) broom::tidy(lm(Petal_Length ~ Petal_Width, e)),
  names = c("term", "estimate", "std.error", "statistic", "p.value"),
  group_by = "Species")
```
The ability to use R packages in Spark is a killer feature, i.e., it expands the capability of Spark to most of the 13,000+ R packages. There are limitations, however. For example, referencing free variables using closures will not work.

```{r}
spark_disconnect(sc)
```