---
title: "Naive Bayes"
author: "Jim Harner"
date: "5/31/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
`sparklyr` requires a `dplyr` compatible back-end to Spark.
```{r}
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(modelr)
library(ggplot2)

library(sparklyr)
# start the sparklyr session
master <- "local"
# master <- "spark://master:7077"
sc <- spark_connect(master)
```

## 6.8 Naive Bayes

### 6.8.1 Basics

Naive Bayes is a classification method that scales well.

#### Bayes Law

Suppose we are testing for a rare disease, where 1% of the population is infected. Now suppose we have a highly sensitive test:

* 99% of sick patients test positive.  
* 99% of healthy patients test negative. 

Given that a patient tests positive, what is the probability that the patient is actually sick?

Suppose we have 10,000 perfectly representative people. That would mean that 100 are sick, and 9,900 are healthy.

From basic probability:
$$
  p(x, y) = p(y\, |\, x)p(x) = p(x\, |\, y)p(y)
$$
Solving for $p(y\, |\, x)$ we get *Bayes's Law*:
$$
  p(y\, |\, x) = \frac{p(x\, |\, y)p(y)}{p(x)}
$$
The denominator term, $p(x)$, can be treated as a *normalization* constant. Set $y$ to to the event "sick" and set $x$ to refer to the event "the test is positive," or "+" for shorthand. Then:  
$$
\begin{align*}
  p(sick\, |\, +) &= \frac{p(+\, |\, sick)p(sick)}{p(+)}
              &= \frac{p(+\, |\, sick)p(sick)}{p(+\, |\, sick)p(sick) + p(+\, |\, well)p(well)}
              &= \frac{0.99 \cdot 0.01}{0.99 \cdot 0.01 + 0.01 \cdot 0.99}
              &= 0.5
\end{align*}
$$
Are you surprised?

#### A spam filter for individual words

We can use Bayes Law to create a spam filter. Let's focus on a single word.
$$
  p(spam\, |\, word)  = \frac{p(word\, |\, spam) p(spam)}{p(word)}
                  = \frac{p(word\, |\, spam) p(spam)}{p(word\, |\, spam) p(spam) 
                    + p(word\, |\, ham) p(ham)},
$$
where $p(ham) = 1 - p(spam$). The right hand side is computable if we have enough labeled data, i.e., if we have training data in which the emails are known to be "spam" or "ham".

#### A Spam Filter That Combines Words: Naive Bayes

We now need to combine words to detect spam. Let: $x = [x_1, x_2, \cdots, x_p]$ be a binary word vector, where $x_j = 1$ or $0$ depending on weather or not the $j^\mbox{th}$ word appears in the email.

Let $c$ denote the event ``is spam.'' The probability model of interest is $p(x\, |\, c)$, i.e., the probability the email's word vector is $x$ given that it is spam. For the $i^{\mbox{th}}$ email:
$$
  p(x_i\, |\, c) = \prod_j \theta_{jc}^{x_{ij}} (1 - \theta_{jc})^{(1 - x_{ij})},
$$
where $\theta_{jc}$ is the probability the $j^{\mbox{th}}$ word is present in spam email. These probabilities can be computed from the training data set on word-by-word basis as above.

This calculation assumes word *independence*. This is why we use the word ``naive'' is used, i.e., certain words tend to appear together, which implies dependence.

By taking logs:
$$
  log(p(x_i\, |\, c)) = \sum_j x_{ij} log(\theta_{jc} / (1 - \theta_{jc})) + \sum_j log(1 - \theta_{jc}).
$$
Let $w_{jc} = log(\theta_{jc} / (1 - \theta_{jc}))$ and $w_{0c} = \sum_j log(1 - \theta_{jc})$, neither of which depend on a given email. Then,
$$
  log(p(x_i\, |\, c)) = \sum_j x_{ij} w_{jc} + w_{0c}.
$$
This is in the form of a *generalized additive model*.

Now that we know how to compute $p(x_i\, |\, c)$, we can use Bayes' Law to compute $p(c\, |\, x_i)$. We must also compute $p(x_i\, |\, h)$, where $h$ denotes ``is ham.'' Basically, we are just counting the words in spam and nonspam emails. If we get more training data, we can easily increment our counts to improve our filter.

#### Laplace Smoothing

In the previous section, $\hat\theta_{jc} = n_{jc}/ n_c$, where $n_{jc}$ denotes the number of times word $j$ appears in a spam email and $n_c$ denotes the number of spam emails. In addition, $\hat{\theta}_{jh} = 1 - \hat{\theta}_{jc}$.

*Laplace Smoothing* refers to the idea of replacing our straight-up estimate of $\hat{θ}_j$ with something a bit fancier:  
$$
  \hat{\theta}_{jc} = \frac{n_{jc} + \alpha}{n_c + \beta},
$$
where $\alpha \ge 0$ and $\beta \ge 0$ are smoothing parameters to avoid having the probability equal 0 or 1. Suppose we denote by `ML` the *maximum likelihood estimate*, and by $D$ the dataset, then we have:
$$
  \theta_{ML} = argmax_{\theta}\  p(D\, |\, \theta_{jc}),
$$
that is for what value of $\theta_{jc}$ were the data $D$ most probable.

If we take the derivative of $log(\theta_{jc}^{n_{jc}} (1 - \theta_{jc})^{n_c - n_{jc}})$ with respect to $\theta_{jc}$, and set it to zero, we get:
$$
  \hat{\theta}_{jc} = n_{jc}/n_c,
$$
which is what we intuitively used before if independence is assumed.

If we put a prior on $\theta_{jc}$, the *a posteriori* distribution is given by:
$$
  p(\theta\, |\, D) =  k \times p(D\, |\, \theta) p(\theta),
$$
where $k$ is a constant and $p(\theta)$ is the prior (the subscripts are dropped below to simplify the notation). The Bayes estimator is the maximum $\theta$ given the data, i.e., the *maximum a posteriori likelihood* (MAP) is given by:
$$
  \theta_{MAP} = argmax_{\theta}\, p(\theta\, |\, D).
$$

If the prior is chosen to be:
$$
  p(\theta) = \theta^{\alpha}(1 - \theta)^{\beta},
$$
then we get the above Laplacian estimate.

How are $\alpha$ and $\beta$ chosen? Generally, we want $\alpha > 0$ and $\beta > 0$ so that the distribution vanishes as 0 and 1. This means that we get few words that never appear or always appear in spam. Likewise we do not want $\alpha$ and $\beta$ to be too large 

$\alpha$ and $\beta$ are called *hyperparameters* or in this case *pseudocounts*. These give you two smoothing knobs. The curse of dimensionality resulting from large feature sets is not a problem for naive Bayes.

The form of the Laplace smoother is often:
$$
  \hat{\theta}_{jc} = \frac{n_{jc} + \alpha}{n_c + \alpha d},
$$
where $\alpha \ge 0$ is the smoothing parameter and $d$ is the number of levels of the classification variable, e.g., $d = 2$ here. This additive smoothing is a *shrinkage estimator*, which shrinks the estimate from $n_{jc}/n_c$ to the uniform probability $1/d$. Often $\alpha = 1$ is used, i.e., *add-one smoothing*.


### 6.8.2 Naive Bayes Wine Quality Example

Read the `winequality-red.csv` file directory into a Spark DataFrame using `spark_red_sdf`.
```{r}
wine_red_sdf <- spark_read_csv(sc, "wine_red_sdf",
                               path = "file:///home/rstudio/rspark-tutorial/m6_supervised_learning/s8_nbayes/wine/winequality-red.csv",
                               delimiter = ";" )
wine_red_tbl <- sdf_register(wine_red_sdf, name = "wine_red_tbl")
```

We split `wine_red_sdf` into a training and a test Spark DataFrame. First, we need to cast `quality` as numeric in order to binarize it with a threshold.
```{r}
wine_red_partition <- wine_red_tbl %>%
  mutate(quality = as.numeric(quality)) %>%
  ft_binarizer(input_col = "quality", output_col = "quality_bin",
               threshold = 5.0) %>%
  sdf_partition(training = 0.7, test = 0.3, seed = 2)
# Create table references
wine_red_train_tbl <- wine_red_partition$training
wine_red_test_tbl <- wine_red_partition$test
```

The default naive bayes classifier has a smoothing parameter of `lambda = 0`.
```
## Specify the model formula
ml_formula <- formula(quality_bin ~ fixed_acidity + volatile_acidity +
                        citric_acid + free_sulfur_dioxide +
                        total_sulfur_dioxide + sulphates + alcohol)
## Naive Bayes
ml_nb <- ml_naive_bayes(wine_red_train_tbl, ml_formula)
summary(ml_nb)
```

We apply Laplace smoothing to the estimates with the commonly used `lambda = 1`.
```
ml_nb <- ml_naive_bayes(wine_red_train_tbl, ml_formula, smoothing = 1)
summary(ml_nb)
```

```{r}
spark_disconnect(sc)
```
